Evaluating the Performance and Energy Efficiency of the COSMO-ART model system

### 2014-04-24 paper_enahpc14/Introduction.tex
gets rid of DOI information in paper.bib

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of   solar   radiation,    thereby   exerting   a   direct   radiative
forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
modification of  cloud properties, cloud fraction  and surface albedo,
causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and    measurement-based    assessments   \citep{Myhre-2009},    large
uncertainties  remain  in   current  estimates  of  aerosol  radiative
forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
Rosenfeld-2013, Sherwood-2013, Stier-2013}.

@@

The extended  COSMO-ART model  is computationally much  more demanding
than COSMO  since a large  number of additional tracers  and processes
have to  be considered.  Thus the model  is currently severely-limited
in  terms   of  applicability  and   expensive  in  terms   of  energy
consumption.    Although   COSMO   has   recently   been   ported   to
GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
High Performance and High  Productivity Computing (HP2C) Initiative to
optimise    it    for     computational    and    energy    efficiency
(\url{http://www.hp2c.ch/}), significant investments  in ART are still
required to take  it to a similar level.   The efficiency of COSMO-ART
is     being    addressed    in     the    EU     Exa2Green    project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

This  work is organized  as follows:  in Sec.~\ref{sec:1}  we describe
related  work,  then in  Sec.~\ref{sec:2},  we  briefly introduce  the
regional atmospheric  model COSMO-ART and specify  its technical setup
related to the investigated performance and energy evaluation methods.
Sec.~\ref{sec:5} describes the HPC  supercomputers used to compute the
baseline.  Sec.~\ref{sec:3}   describes  different  power  measurement
equipment  used for  conducting this  benchmarking study  and presents
energy-aware techniques employed to  determine the energy footprint of
the   considered  application.   Sec.~\ref{sec:4}   gathers  benchmark
results  from  the  Exa2Green   project  and  highlights  areas  where
improvements   will   be  necessary   for   the  subsequent   baseline
refactoring.   Finally,  we  conclude  with  practical  guidelines  to
identifying energy saving opportunities  and give an outlook to future
research in Sec.~\ref{concl}.

>>>>

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of solar radiation, thereby exerting a direct radiative forcing; or by
the  modification  of cloud  properties,  cloud  fraction and  surface
albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and  measurement-based  assessments,  large  uncertainties  remain  in
current  estimates  of  aerosol radiative  forcing  \citep{Myhre-2013,
IPCC-2013,  Lee-2013,   Randles-2013,  Rosenfeld-2013,  Sherwood-2013,
Stier-2013}.

@@

The extended atmospheric model  COSMO-ART is computationally much more
demanding than  COSMO since a  large number of additional  tracers and
processes  have  to  be  considered.   Thus  the  model  is  currently
severely-limited in  terms of applicability and expensive  in terms of
energy  consumption.   Although  COSMO  has recently  been  ported  to
GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
High    Performance   and    High   Productivity    Computing   (HP2C)
Initiative \citep{HP2C},  to optimise it for  computational and energy
efficiency, significant investments in  ART are still required to take
it to a similar level.  The efficiency of COSMO-ART is being addressed
in the  EU Exa2Green project \citep{EXA2GREEN} to  deliver a prototype
code, which  provides an energy efficiency  of at least  five times of
the baseline value.  Such  an implementation would allow the community
to investigate critical questions at higher resolution and over longer
periods, at reduced cost to the environment.

This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
introduce  COSMO-ART and specify  its technical  setup related  to the
investigated   performance   and   energy  evaluation   methods.    In
Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking study.  Sec.~\ref{sec:4}  presents performance and energy
requirements  of the  baseline on  these architectures  and highlights
areas where improvements will be necessary for the subsequent baseline
refactoring.   Finally,  we  conclude  with  practical  guidelines  to
identifying energy saving opportunities  and give an outlook to future
research in Sec.~\ref{concl}.
